Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-27 survey/survey.tex
some integration in sec3

\begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
\label{def:primitives-langRev}

\[\begin{array}{rrcll}
 \mathit{zeroe} :& 0 + b &\leftrightarrow & b &: \mathit{zeroi} \\
 \mathit{swap}^{+} :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \mathit{swap}^{+} \\
 \mathit{assocl}^{+} :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \mathit{assocr}^{+} \\
 \mathit{unite} :& 1 \times b &\leftrightarrow & b &: \mathit{uniti} \\
 \mathit{swap}^{\times } :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \mathit{swap}^{\times } \\
 \mathit{assocl}^{\times } :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \mathit{assocr}^{\times } \\
 \mathit{distrib }_{0} :& 0 \times b &\leftrightarrow & 0 &: \mathit{factor }_{0} \\
 \mathit{distrib } :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \mathit{factor } \\
 \end{array}\]
%subcode source isomorphisms.tex:427

Each line of this table is to be read as the definition of one or two
operators. For example, corresponding to the \textit{identity of
  \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
\ensuremath{\mathit{unite}:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
\ensuremath{\mathit{uniti}:b\leftrightarrow 1\times b} (reading the isomorphism from right to
left). These operators are inverses of each other.  Each of the two
cases of commutativity defines one operator that is its own inverse.

Having defined primitive operators, we need some means of composing
them. We construct the composition combinators out of the closure
conditions for isomorphisms.

\begin{definition}{Composition in \ensuremath{\Pi } }
$$
\infer{ \mathit{id} : b \leftrightarrow b}{
	 ~
}
\quad
\infer{ \mathit{sym} ~c : b_2 \leftrightarrow b_1}{
	 c : b_1 \leftrightarrow b_2
}
\quad
\infer{ c_1\fatsemi c_2 : b_1 \leftrightarrow b_3}{
	 c_1 : b_1 \leftrightarrow b_2
	&
	 c_2 : b_2 \leftrightarrow b_3
}
$$
$$
\infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
	 c_1 : b_1 \leftrightarrow b_3
	&
	 c_2 : b_2 \leftrightarrow b_4
}
\quad
\infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
	 c_1 : b_1 \leftrightarrow b_3
	&
	 c_2 : b_2 \leftrightarrow b_4
}
$$
%subcode source isomorphisms.tex:462
\end{definition}

Thus we have program constructs that witness reflexivity \ensuremath{\mathit{id}},
symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\fatsemi} and two parallel
composition combinators, one for sums~\ensuremath{+} and one for
pairs~\ensuremath{\times}.

\begin{definition}{(Syntax of \ensuremath{\Pi })}
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition.
\[\begin{array}{rclr}
 \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
 \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
 \\
 \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
 \mathit{isomorphisms}, \mathit{iso} &::=& \mathit{zeroe} ~|~ \mathit{zeroi} ~|~ \mathit{swap}^{+} ~|~ \mathit{assocl}^{+} ~|~ \mathit{assocr}^{+} \\
 & ~|~ & \mathit{unite} ~|~ \mathit{uniti} ~|~ \mathit{swap}^{\times } ~|~ \mathit{assocl}^{\times } ~|~ \mathit{assocr}^{\times } \\
 & ~|~ & \mathit{distrib }_{0} ~|~ \mathit{factor }_{0} ~|~ \mathit{distrib } ~|~ \mathit{factor } \\
 \mathit{combinators}, c &::=& \mathit{iso} ~|~ \mathit{id} ~|~ \mathit{sym} ~c ~|~ c \fatsemi c ~|~ c + c ~|~ c \times c \\
 \end{array}\]
%subcode source isomorphisms.tex:483
\end{definition}

@@

\[\begin{array}{rlcl}
 \mathit{zeroe} & (\mathit{right} ~v) &\mapsto & v \\
 \mathit{zeroi} & v &\mapsto & \mathit{right} ~v \\
 \mathit{swap}^{+} & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
 \mathit{swap}^{+} & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
 \mathit{assocl}^{+} & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
 \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
 \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
 \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
 \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
 \mathit{assocr}^{+} & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:504

@@

\[\begin{array}{rlcl}
 \mathit{unite} & ((), v) &\mapsto & v \\
 \mathit{uniti} & v &\mapsto & ((), v) \\
 \mathit{swap}^{\times } & (v_1, v_2) &\mapsto & (v_2, v_1) \\
 \mathit{assocl}^{\times } & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
 \mathit{assocr}^{\times } & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
 \end{array}\]
%subcode source isomorphisms.tex:514

@@

\[\begin{array}{rlcl}
 \mathit{distrib } & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
 \mathit{distrib } & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
 \mathit{factor } & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
 \mathit{factor } & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:523

@@

$$
\infer{ \mathit{id} ~v \mapsto v}{
	 ~
}
\quad
\infer{ (\mathit{sym} ~c) ~v_1 \mapsto v_2}{
	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
}
\quad
\infer{ (c_1\fatsemi c_2) ~v_1 \mapsto v_2}{
	 c_1 ~v_1 \mapsto v
	&
	 c_2 ~v \mapsto v_2
}
$$
$$
\infer{ (c_1 + c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
	 c_1 ~v_1 \mapsto v_2
}
\quad
\infer{ (c_1 + c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
	 c_2 ~v_1 \mapsto v_2
}
$$
$$
\infer{ (c_1 \times c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
	 c_1 ~v_1 \mapsto v_3
	&
	 c_2 ~v_2 \mapsto v_4
}
$$
%subcode source isomorphisms.tex:546

@@

Since there are no values that have the type \ensuremath{0}, the reductions for
the combinators \ensuremath{\mathit{zeroe}}, \ensuremath{\mathit{zeroi}}, \ensuremath{\mathit{distrib }_{0}} and \ensuremath{\mathit{factor }_{0}}
omit the impossible cases. Also, the semantics of \ensuremath{\mathit{sym} ~c} are
defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
which we formalize below.

@@

  \item \ensuremath{(c_1 \times c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \times c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1 + c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } + c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1\fatsemi c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \fatsemi c_1^\mathit{^{\dagger} }}. (Note that the
    order of combinators has been reversed).

@@

\[\begin{array}{rclr}
 \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \mathit{id} \\
 \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \mathit{id} \\
 \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \mathit{id} + \lbrack\!\lbrack b \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{+} \fatsemi \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack + \mathit{id}) \fatsemi \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{\times } \fatsemi \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{distrib } \fatsemi \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
 \end{array}\]
%subcode source isomorphisms.tex:598

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Graphical Language for \ensuremath{\Pi } }

@@

\begin{itemize}
\item
The simplest sort of diagram is the \ensuremath{\mathit{id} : b \leftrightarrow b} combinator which
is simply represented as a wire labeled by its type \ensuremath{b}. In more
complex diagrams, if the type of a wire is obvious from the context,
it may be omitted.

@@

\item Combinators can be composed in series (\ensuremath{c_1 \fatsemi c_2}) or
  parallel. Sequential (series) composition corresponds to connecting
  the output of one combinator to the input of the next.

@@

\ensuremath{c = \mathit{swap}^{\times } \fatsemi \mathit{distrib } \fatsemi (\mathit{unite} + \mathit{unite})}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Expressing Programs with \ensuremath{\Pi } }
\label{sec:langRev-examples}

@@

\ensuremath{\mathit{not} = \mathit{swap}^{+}}

@@

\ensuremath{\mathit{reverse} = \mathit{swap}^{\times } \fatsemi (\mathit{swap}^{\times } \times \mathit{id})~ \fatsemi \mathit{assocr}^{\times }}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \mathit{swap}^{\times } & ((v_2, v_3), v_1) \\
 \mathit{swap}^{\times } \times \mathit{id} & ((v_3, v_2), v_1) \\
 \mathit{assocr}^{\times } & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

\noindent On the first line is the initial value. On each subsequent
line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
results from applying this combinator to the value on the previous
line. For example, \ensuremath{\mathit{swap}^{\times }} transforms \ensuremath{(v_1, (v_2, v_3))} to
\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
the bits in reverse order.

@@

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\mathit{distrib }} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be \ensuremath{\mathit{bool}} and apply either
\ensuremath{c_1:b\leftrightarrow b} or \ensuremath{c_2:b\leftrightarrow b} to the value \ensuremath{v}, then we essentially
have an `if' expression.

@@

\ensuremath{\mathit{if}_{c_1,c_2} = \mathit{distrib } \fatsemi ((\mathit{id} \times c_1) + (\mathit{id} \times c_2)) \fatsemi \mathit{factor }}

@@

The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
by the distribute operator \ensuremath{\mathit{distrib }}, which converts it into a value of
type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was \ensuremath{\mathit{left} ~()}),
the combinator~\ensuremath{c_1} is applied to the value of type~\ensuremath{b}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b} through the combinator \ensuremath{c_2}.

@@

\label{translation}
\noindent
Every simple isomorphism of Theseus can be translated to \ensuremath{\Pi^{o}} as follows:
\begin{enumerate}
\item For every Theseus map \ctr{c : a} $\leftrightarrow$ \ctr{b}, we can
  define an intermediate type \ctr{ti} which is the sum of pairs of all the
  variables that occur in the LHS patterns. Since the LHS and RHS patterns
  contain the same type variables, \ctr{ti} is unique for a given \ctr{c} (up
  to ordering). For example, in \ctr{treeUnwind} the first clause has
  variables \lstinline{t1 : Tree} and \lstinline{t2 : Tree}, the second and
  third clauses have no variables and the fourth clause has the variable
  \lstinline{n : Nat}. The expected type \ctr{ti} is therefore
  \lstinline{Tree * Tree + (1 + (1 + Nat))}, where we insert the type
  {\scriptsize{1}} as a placeholder for the clauses that did not have type
  variables.
\item By construction \ctr{ti} is isomorphic to \ctr{a} and \ctr{b}. Hence,
  maps \ctr{c1 : a} $\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti}
  $\leftrightarrow$ \ctr{b} must be expressible in \ensuremath{\Pi^{o}}. (Both the above
  isomorphisms can be constructed by systematically unfolding and
  distributing \ctr{a} and \ctr{b} until \ctr{ti} is obtained.  The exact
  details of such an expansion can be found in Sec.~3 of our previous
  paper~\cite{rc2012} and are skipped in this paper.) Once values of type
  \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
  translation of \ctr{c} is given by \ctr{c1} $\fatsemi$ \ctr{c2}.
\end{enumerate}

@@

\noindent
Compiling programs with labels to \ensuremath{\Pi^{o}} is a straightforward extension of
compiling label-free programs. For any Theseus map \ctr{c : a}
$\leftrightarrow$ \ctr{b} that uses an iteration label \kw{label} \ctr{: lab}
we construct an intermediate type \ctr{ti} corresponding to the pattern
clauses as before. We can then construct the isomorphisms \ctr{c1 : lab + a}
$\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti} $\leftrightarrow$ \ctr{lab + b}
since every LHS pattern clause corresponds to either \ctr{a} (the input) or
\ctr{lab} (the label) and every RHS clause corresponds to either \ctr{b} (the
output) or \ctr{lab} (the label). The resulting \ctr{c1} $\fatsemi$ \ctr{c2}
has type \lstinline{lab+a} $\leftrightarrow$ \lstinline{lab+b} and the
required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\fatsemi$
\ctr{c2} \ensuremath{)} \ctr{: a} $\leftrightarrow$ \ctr{b}. This extends to multiple
by labels by constructing \lstinline{c1 : (lab1 + lab2 + ...) + a}
$\leftrightarrow$ \ctr{ti} and \lstinline{c2 : ti} $\leftrightarrow$
\lstinline{(lab1 + lab2 + ...) + b} to include the types of the additional
labels.

>>>>

\begin{definition}{(Syntax of \ensuremath{\Pi })}
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition.
\[\begin{array}{rclr}
 \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
 \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
 \\
 \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
 \mathit{isomorphisms}, \mathit{iso} &::=& \identlp ~|~ \identrp ~|~ \swapp ~|~ \assoclp ~|~ \assocrp \\
 & ~|~ & \identlt ~|~ \identrt ~|~ \swapt ~|~ \assoclt ~|~ \assocrt \\
 & ~|~ & \absorbr ~|~ \factorzl ~|~ \dist ~|~ \factor \\
 \mathit{combinators}, c &::=& \mathit{iso} ~|~ \idc ~|~ \ !~c ~|~ c \odot c ~|~ c \oplus c ~|~ c \otimes c \\
 \end{array}\]
%subcode source isomorphisms.tex:483
\end{definition}

@@

\[\begin{array}{rlcl}
 \identlp & (\mathit{right} ~v) &\mapsto & v \\
 \identrp & v &\mapsto & \mathit{right} ~v \\
 \swapp & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
 \swapp & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
 \assoclp & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
 \assoclp & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
 \assoclp & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
 \assocrp & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
 \assocrp & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
 \assocrp & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:504

@@

\[\begin{array}{rlcl}
 \identlt & ((), v) &\mapsto & v \\
 \identrt & v &\mapsto & ((), v) \\
 \swapt & (v_1, v_2) &\mapsto & (v_2, v_1) \\
 \assoclt & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
 \assocrt & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
 \end{array}\]
%subcode source isomorphisms.tex:514

@@

\[\begin{array}{rlcl}
 \dist & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
 \dist & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
 \factor & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
 \factor & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:523

@@

$$
\infer{ \idc ~v \mapsto v}{
	 ~
}
\quad
\infer{ (\ !~c) ~v_1 \mapsto v_2}{
	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
}
\quad
\infer{ (c_1\odot c_2) ~v_1 \mapsto v_2}{
	 c_1 ~v_1 \mapsto v
	&
	 c_2 ~v \mapsto v_2
}
$$
$$
\infer{ (c_1 \oplus c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
	 c_1 ~v_1 \mapsto v_2
}
\quad
\infer{ (c_1 \oplus c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
	 c_2 ~v_1 \mapsto v_2
}
$$
$$
\infer{ (c_1 \otimes c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
	 c_1 ~v_1 \mapsto v_3
	&
	 c_2 ~v_2 \mapsto v_4
}
$$
%subcode source isomorphisms.tex:546

@@

Since there are no values that have the type \ensuremath{0}, the reductions for
the combinators \ensuremath{\identlp}, \ensuremath{\identrp}, \ensuremath{\absorbr} and \ensuremath{\factorzl}
omit the impossible cases. Also, the semantics of \ensuremath{\ !~c} are
defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
which we formalize below.

@@

  \item \ensuremath{(c_1 \otimes c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \otimes c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1 \oplus c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \oplus c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1\odot c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \odot c_1^\mathit{^{\dagger} }}. (Note that the
    order of combinators has been reversed).

@@

\[\begin{array}{rclr}
 \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \idc \\
 \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \idc \\
 \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \idc \oplus \lbrack\!\lbrack b \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \assocrp \odot \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack \oplus \idc) \odot \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \assocrt \odot \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \dist \odot \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
 \end{array}\]
%subcode source isomorphisms.tex:598

@@

@@

\begin{itemize}
\item
The simplest sort of diagram is the \ensuremath{\idc : b \leftrightarrow b} combinator which
is simply represented as a wire labeled by its type \ensuremath{b}. In more
complex diagrams, if the type of a wire is obvious from the context,
it may be omitted.

@@

\item Combinators can be composed in series (\ensuremath{c_1 \odot c_2}) or
  parallel. Sequential (series) composition corresponds to connecting
  the output of one combinator to the input of the next.

@@

\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}

@@

% \begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
% \label{def:primitives-langRev}

% \[\begin{array}{rrcll}
%  \identlp :& 0 + b &\leftrightarrow & b &: \identrp \\
%  \swapp :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \swapp \\
%  \assoclp :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \assocrp \\
%  \identlt :& 1 \times b &\leftrightarrow & b &: \identrt \\
%  \swapt :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \swapt \\
%  \assoclt :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \assocrt \\
%  \absorbr :& 0 \times b &\leftrightarrow & 0 &: \factorzl \\
%  \dist :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \factor \\
%  \end{array}\]
% %subcode source isomorphisms.tex:427

% \end{definition}

% Each line of this table is to be read as the definition of one or two
% operators. For example, corresponding to the \textit{identity of
%   \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
% \ensuremath{\identlt:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
% \ensuremath{\identrt:b\leftrightarrow 1\times b} (reading the isomorphism from right to
% left). These operators are inverses of each other.  Each of the two
% cases of commutativity defines one operator that is its own inverse.

% Having defined primitive operators, we need some means of composing
% them. We construct the composition combinators out of the closure
% conditions for isomorphisms.

% \begin{definition}{Composition in \ensuremath{\Pi } }
% $$
% \infer{ \idc : b \leftrightarrow b}{
% 	 ~
% }
% \quad
% \infer{ \ !~c : b_2 \leftrightarrow b_1}{
% 	 c : b_1 \leftrightarrow b_2
% }
% \quad
% \infer{ c_1\odot c_2 : b_1 \leftrightarrow b_3}{
% 	 c_1 : b_1 \leftrightarrow b_2
% 	&
% 	 c_2 : b_2 \leftrightarrow b_3
% }
% $$
% $$
% \infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
% 	 c_1 : b_1 \leftrightarrow b_3
% 	&
% 	 c_2 : b_2 \leftrightarrow b_4
% }
% \quad
% \infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
% 	 c_1 : b_1 \leftrightarrow b_3
% 	&
% 	 c_2 : b_2 \leftrightarrow b_4
% }
% $$
% %subcode source isomorphisms.tex:462
% \end{definition}

% Thus we have program constructs that witness reflexivity \ensuremath{\idc},
% symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\odot} and two parallel
% composition combinators, one for sums~\ensuremath{+} and one for
% pairs~\ensuremath{\times}.

\label{sec:langRev-examples}

@@

\ensuremath{\mathit{not} = \swapp}

@@

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt \times \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \times \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

\noindent On the first line is the initial value. On each subsequent
line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
results from applying this combinator to the value on the previous
line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
the bits in reverse order.

@@

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be \ensuremath{\mathit{bool}} and apply either
\ensuremath{c_1:b\leftrightarrow b} or \ensuremath{c_2:b\leftrightarrow b} to the value \ensuremath{v}, then we essentially
have an `if' expression.

@@

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc \times c_1) + (\idc \times c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
by the distribute operator \ensuremath{\dist}, which converts it into a value of
type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was \ensuremath{\mathit{left} ~()}),
the combinator~\ensuremath{c_1} is applied to the value of type~\ensuremath{b}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b} through the combinator \ensuremath{c_2}.

@@

\label{translation}
\noindent
Every simple isomorphism of Theseus can be translated to \ensuremath{\Pi^{o}} as follows:
\begin{enumerate}
\item For every Theseus map \ctr{c : a} $\leftrightarrow$ \ctr{b}, we can
  define an intermediate type \ctr{ti} which is the sum of pairs of all the
  variables that occur in the LHS patterns. Since the LHS and RHS patterns
  contain the same type variables, \ctr{ti} is unique for a given \ctr{c} (up
  to ordering). For example, in \ctr{treeUnwind} the first clause has
  variables \lstinline{t1 : Tree} and \lstinline{t2 : Tree}, the second and
  third clauses have no variables and the fourth clause has the variable
  \lstinline{n : Nat}. The expected type \ctr{ti} is therefore
  \lstinline{Tree * Tree + (1 + (1 + Nat))}, where we insert the type
  {\scriptsize{1}} as a placeholder for the clauses that did not have type
  variables.
\item By construction \ctr{ti} is isomorphic to \ctr{a} and \ctr{b}. Hence,
  maps \ctr{c1 : a} $\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti}
  $\leftrightarrow$ \ctr{b} must be expressible in \ensuremath{\Pi^{o}}. (Both the above
  isomorphisms can be constructed by systematically unfolding and
  distributing \ctr{a} and \ctr{b} until \ctr{ti} is obtained.  The exact
  details of such an expansion can be found in Sec.~3 of our previous
  paper~\cite{rc2012} and are skipped in this paper.) Once values of type
  \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
  translation of \ctr{c} is given by \ctr{c1} $\odot$ \ctr{c2}.
\end{enumerate}

@@

\noindent
Compiling programs with labels to \ensuremath{\Pi^{o}} is a straightforward extension of
compiling label-free programs. For any Theseus map \ctr{c : a}
$\leftrightarrow$ \ctr{b} that uses an iteration label \kw{label} \ctr{: lab}
we construct an intermediate type \ctr{ti} corresponding to the pattern
clauses as before. We can then construct the isomorphisms \ctr{c1 : lab + a}
$\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti} $\leftrightarrow$ \ctr{lab + b}
since every LHS pattern clause corresponds to either \ctr{a} (the input) or
\ctr{lab} (the label) and every RHS clause corresponds to either \ctr{b} (the
output) or \ctr{lab} (the label). The resulting \ctr{c1} $\odot$ \ctr{c2}
has type \lstinline{lab+a} $\leftrightarrow$ \lstinline{lab+b} and the
required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\odot$
\ctr{c2} \ensuremath{)} \ctr{: a} $\leftrightarrow$ \ctr{b}. This extends to multiple
by labels by constructing \lstinline{c1 : (lab1 + lab2 + ...) + a}
$\leftrightarrow$ \ctr{ti} and \lstinline{c2 : ti} $\leftrightarrow$
\lstinline{(lab1 + lab2 + ...) + b} to include the types of the additional
labels.
